Model Averaging

### 2016-06-09 finalwriteup/draft.rmd
report writing profit curve

@@

* tried model averaging - comparison to other methods.

@@

* compare to past research. problems with past research and does this solve any of those?
    * poor user uptake - future work?
    * lack of trust - nearest neighbour must be presented with results - case bsaed reasoning
        * results from boosted tree shows partial dependence plots - insight.
    * never been applied to a consultant within construction industry. has been applied lots to software
        * still has the potential to help with the way the industry is structured - hinges on user uptake.

* this shows the profitability can be predicted with some success, but real question is, is it financially viable?
* depending what the bottom line plot shows - yes or no.

>>>>

### Bottom Line Analysis

The first research question, whether machine learning and statistical techniques can predict project profitability based on historic data, has been answered. The measure of profitability could not be modelled but simplifying the profitability prediction to a binary (profit/loss) model was successful. To answer the second research question, whether the model would improve the business' profitability, a bottom line analysis was performed.

The models output a 'probability' between 0 and 1 that each project will be a loss-making job (where probability = 1). This leaves the question, at what probability would a decision-maker round the probability up to 1 or down to 0? And what business decision would then be made? To find the rounding threshold, an experimental business-scenario was tested. At a given threshold, all projects calculated above the threshold were considered too risky, and were rejected. All profits and losses from these projects were forfeited, and the profits from the remaining jobs (below the threshold) were summed to give a total profit.

Using this logic, if the threshold was 0, all projects were rejected and total profit would equal 0. If the threshold was 1, all projects were accepted and the total profits would be the same as historical figures. Probability results from the two blended models as well as three individual models were tested and because the models require a training and testing set, the process was repeated 100 times with 20 x 5 fold cross validation models to achieve a statistical power of 0.8. The plot below illustrates the results of each model at each threshold with a local polynomial regression line of best fit (loess) and a 95% confidence interval around the mean.

```{r profit_curve, fig.depth = 10}
finance.plot = readRDS("report_data/plot_finance.rds" )

ggplot(finance.plot, aes(x= threshold, y= profit.ratio, colour = method)) + 
        geom_point(alpha = 0.1, shape = 16) + 
        stat_smooth() +
        scale_y_continuous(breaks = seq(0,110, by = 10)) + 
        scale_x_continuous(breaks = seq(0,1, by = .1)) + 
        geom_hline(yintercept = 100, colour = "darkgray", linetype = 2) + 
        geom_hline(yintercept = 0, colour = "darkgray", linetype = 1) + 
        theme(legend.position = "bottom") + 
        scale_colour_discrete(labels = c("average" = "Blend - Average",
                                       "orig.boost"= "Blend - Log Reg",
                                       "orig.log" = "Logistic Regression",
                                       "orig.rf" = "Random Forest",
                                       "simp.log" = "Boosted Trees"))+
        guides(col = guide_legend(nrow = 2)) +
        labs(x = "Probability Threshold", 
             y = "Ratio of Model Profit to Historical Profit (%)",
             title = "Profit Curve for Blended and Individual Models")

The two blended models outperformed the individual models as shown in this table describing each models' optimal thresholds:

```{r profit_table}

finance.plot$method = as.factor(finance.plot$method)
# want mean and sd for each method at each threshold

fin.summary = finance.plot %>% group_by(threshold, method) %>% 
        summarise_each(funs(mean_profit_ratio = round(mean(.), 1), 
                            sd_prof_ratio = round(sd(.), 1)),
                       profit.ratio)

#for each method find max mean

best.fin = fin.summary %>% group_by(method) %>% 
        filter(mean_profit_ratio == max(mean_profit_ratio)) %>% 
        arrange(desc(mean_profit_ratio))
kable(best.fin)

best.mean = best.fin %>% filter(method == "average") %>% .$mean_profit_ratio
best.sd = best.fin %>% filter(method == "average") %>% .$sd_prof_ratio

The average of the three individual models performed best with the highest mean profit ratio at `r best.mean`% and the narrowed standard deviation at `r best.sd`%. This means that for the blended average model, if all jobs above the probability threshold 0.55 were rejected, the profits would increase on average by `r best.mean - 100`% in comparison to historical profit data.

For clarity, the plot for the average blended model is shown below:

```{r av_profit_curve}
ggplot(finance.plot %>% filter(method == "average"), 
       aes(x= threshold, y= profit.ratio, colour = method)) + 
        geom_point(alpha = 0.1, shape = 16) + 
        stat_smooth() +
        scale_y_continuous(breaks = seq(0,110, by = 10)) + 
        scale_x_continuous(breaks = seq(0,1, by = .1)) + 
        geom_hline(yintercept = 100, colour = "darkgray", linetype = 2) + 
        geom_hline(yintercept = 0, colour = "darkgray", linetype = 1) + 
        theme(legend.position = "none") +
        labs(x = "Probability Threshold", 
             y = "Ratio of Model Profit to Historical Profit (%)",
             title = "Profit Curve for Blended Average Model")

@@

@@

comments on your results;
* blended models significantly (?) outperformed individual models
*

explains what your results mean;

* If the company's profits for one year were $1,000,000, the mean yearly increased profit would be `r (best.mean-100)/100*1000000` by using the best model and rejecting all jobs with a probability higher than `r best.fin %>% filter(method == 'average') %>% .$threshold`.

interprets your results in a wider context; indicates which results were expected or unexpected;
it was expected that it would show financial benefits and the results look reasonable - not particuarly high or low.
it was also expected that blended model would outperform the single models. Interesting that simple model averaging ended up signficantly (?) outperformiing blended lgoistic regression model in profit curve.

provides explanations for unexpected results.
why did it end up outperforming? no idea.

The Discussion should also relate your specific results to previous research or theory. 
* problems with past research and does this solve any of those?
    * poor user uptake - future work?
    * lack of trust - nearest neighbour must be presented with results - case bsaed reasoning
    * results from boosted tree shows partial dependence plots - insight.
    * never been applied to a consultant within construction industry. has been applied lots to software
    * still has the potential to help with the way the industry is structured - hinges on user uptake.

You should point out what the limitations were of your study, and note any questions that remain unanswered. 
* not sure how easy number of users is to predict - could do a study
* could do a study on trust levels

The Discussion CAN also include Conclusions/Future Research.
could be a viable business move worth integrating

### 2016-06-09 financial.rmd
report writing profit curve

```{r_loadpackages, echo=FALSE, include=FALSE}
library("plyr",lib = 'C:/Progra~1/R/R-3.2.1/library')
library("dplyr",lib = 'C:/Progra~1/R/R-3.2.1/library')
library('magrittr',lib='C:/Progra~1/R/R-3.2.1/library')
library('reshape2',lib='C:/Progra~1/R/R-3.2.1/library')
library('ggplot2',lib='C:/Progra~1/R/R-3.2.1/library')
library('gbm', lib = 'C:/Progra~1/R/R-3.2.2/library')
library('randomForest', lib = 'C:/Progra~1/R/R-3.2.2/library')
library('caret', lib = 'C:/Progra~1/R/R-3.2.2/library')
library('pROC',lib='C:/Progra~1/R/R-3.2.2/library')

@@

```{r, echo=FALSE}
test.pred <- read.csv('C:/Users/n9232371/Documents/Consultbusiness/data/test_review.csv')
all7d <- read.csv('C:/Users/n9232371/Documents/Consultbusiness/data/all7d.csv')
test.pred<- merge(test.pred, all7d %>% select(mlsto, balance.mlsto), by = "mlsto")

@@

for(i in 1:length(threshold)){
        prof<- data.frame("balance" = test.pred$balance.mlsto, 
                  "pred" = test.pred$pred,
                  "ans" = test.pred$b.rpdol)
        #pred = probabilities
        prof$conf <- rep(0, nrow(prof))
        prof$conf <- ifelse(prof$pred >= threshold[i] & prof$ans == 1, 'TP', prof$conf)
        prof$conf <- ifelse(prof$pred >= threshold[i] & prof$ans == 0, 'FP', prof$conf)
        prof$profit<- rep(0, nrow(prof))
        prof$profit<- ifelse(prof$conf %in% "TP", -prof$balance, prof$profit)
        prof$profit<- ifelse(prof$conf %in% "FP", -prof$balance, prof$profit)

@@

>>>>

```{r loadpackages, echo=FALSE, include=FALSE}

library("plyr")
library("dplyr")
library('magrittr')
library('reshape2')
library('ggplot2')
library('gbm')
library('randomForest')
library('caret')
library('pROC')
library('testthat')
library('purrr')
library(tidyr)
library(broom)

@@

test.pred <- read.csv('C:/Users/n9232371/Documents/Consultbusiness/data/test_review.csv')
all7d <- read.csv('C:/Users/n9232371/Documents/Consultbusiness/data/all7d.csv')
test.pred<- merge(test.pred, all7d %>% select(mlsto, balance.mlsto), by = "mlsto")

@@

for(i in 1:length(threshold)){
        prof<- data.frame("balance" = test.pred$balance.mlsto, 
                  "pred" = test.pred$pred,
                  "ans" = test.pred$b.rpdol)
        #pred = probabilities
        prof$conf <- rep(0, nrow(prof))
        prof$conf <- ifelse(prof$pred >= threshold[i] & prof$ans == 1, 'TP', prof$conf)
        prof$conf <- ifelse(prof$pred >= threshold[i] & prof$ans == 0, 'FP', prof$conf)
        # if model marks it as a positive (greater than threshold), eliminate any balance
        prof$profit<- rep(0, nrow(prof))
        prof$profit<- ifelse(prof$conf %in% "TP", -prof$balance, prof$profit)
        prof$profit<- ifelse(prof$conf %in% "FP", -prof$balance, prof$profit)

@@

Need a new function for final analysis! comparing two blending methods
Can i get rid of boosted tree model? that would speed things up a lot

Need to try simple average and log reg without boosted trees. Reduces spread but doesnt improve mean. i'd like to keep the wide spread because you reach greater heights (and lower lows)

```{r}
if(.Platform$OS.type == 'windows'){
        all10mice = read.csv('C:/Users/n9232371/OneDrive/shared files/Bligh Tanner/masters/data/all10mice_May16.csv')[,-1]
        source("C:/Users/n9232371/Documents/github/consult/functions.R")
} else{
        all10mice = read.csv('~/OneDrive/shared files/Bligh Tanner/masters/data/all10mice_May16.csv')[,-1]
        source("/Users/yam/Documents/github/consult/functions.R")

profit <- function(df = all10mice, method = c("average","simp.log","orig.log","orig.rf","orig.boost"), 
                   seed = 300, 
                   k.folds = 5, # number of folds in initial model.blend generation
                   max.iter = 5, #max number of iterations from initial model.blend generation
                   k.folds.2 = 5, # number of folds for second step blended model development
                   tot.tests = 50 # total number of results to create
                   ){

        length.k.folds = ceiling(tot.tests/k.folds.2)*k.folds.2

        set.seed(seed)

        # testthat check
        expect_equal(levels(df$f.rpdol), c("profit", "loss"))

        #loop through however many versions of df.blend are required
        compile.results = NULL

        for(m in 1:ceiling(tot.tests/max.iter)){

        print(
                paste("m =", m, "of", ceiling(tot.tests/max.iter),"\n", sep = ' ') %>% cat()
        )

        # create complete set
        df.blend = assemble(methods = c("log","rf","boost"), df.assem = df, x.folds.assem = 5,
                     manual.seed.assem = sample(1:1000,1))

        df.blend = mutate(df.blend, norm.log = sign(log) * abs(log)^(1/3))
        df.blend = mutate(df.blend, norm.rf = rf^0.5)
        df.blend = mutate(df.blend, norm.boost = log(boost))

        saveit <- iter.auc(mult = max.iter/k.folds.2, x.folds.iter = k.folds.2,
                           methods = c("orig.log","average","simp.log", "orig.rf","orig.boost"),
                           df.iter = df.blend)

        dontknow = map(saveit, function(x) spread(x, iteration, pred))

        if(m == 1){
                compile.results = dontknow
        } else {
                compile.results = map2(compile.results, dontknow, function(x,y) merge(x, y, by = 'index'))
        }

        # saveRDS(compile.results, "C:/Users/n9232371/Documents/github/consult/finalwriteup/report_data/finance_results.rds")
        # compile.results <- readRDS("C:/Users/n9232371/Documents/github/consult/finalwriteup/report_data/finance_results.rds")

        ####
        # now have a suite of predictions - merge with original all10.blend$balance.mlsto
        df.blend$index = rownames(df.blend) %>% as.numeric
        # load balances
        all6 <-read.csv('C:/Users/n9232371/OneDrive/shared files/Bligh Tanner/masters/data/all6.csv')
        all6 <- all6 %>% select(mlsto, balance.mlsto) %>% distinct()
        all6$mlsto <- as.character(all6$mlsto)
        df.blend = left_join(df.blend, all6)
        compile.results <- map(compile.results, function(x) 
                left_join(x, df.blend %>% select(index, balance.mlsto, b.rpdol)))

        #create vector of profit points
        threshold = seq(0, 1, by = .05)

        # set up list of blank dataframes for profit calcs for each iteration: prof.results
        colnames = names(compile.results[[1]])[grepl('iter', names(compile.results[[1]]))]
        new.df = matrix(NA, nrow = length(threshold), ncol = ceiling(tot.tests/max.iter) + 1) %>% as.data.frame
        colnames(new.df) <- c('threshold', colnames)
        new.df$threshold = threshold 
        prof.results = list(new.df, new.df, new.df, new.df, new.df)
        names(prof.results) = names(compile.results)

        # try eliminating outliers of balance over $200,000
        # comp.results.2 <- lapply(compile.results, function(x) x = x %>% filter(abs(balance.mlsto)<200000))

        prof.results <- map2(compile.results, prof.results, function(x,y) {

                for(n in 2:ncol(y)){

                        for (p in 1:length(threshold)) {

                        # sum the balance.mlsto of all jobs that fall below the cut off threshold
                        # this equals the total profits (or losses!) across all jobs that were completed
                        # given that you reject any job over the threshold

                        prof <- x[x[,n] < y$threshold[p], 'balance.mlsto']

                        y[p, n] = sum(prof)

                return(y)

        })

        prof.results <- map2(prof.results, names(prof.results), function(x,y) {
                x$method = y
                return(x)})

        assign(prof.results, "prof.results", envir = .GlobalEnv)

# gather results for plotting
plot.results = rbind_all(prof.results)
plot.results = gather(plot.results, key = iteration, value = profit, -method, - threshold)

# censor actual profit amount as fraction of historical profit level
orig.prof = plot.results %>% filter(threshold == 1) %>% .$profit %>% .[1]
plot.results = mutate(plot.results, profit.ratio = (profit/orig.prof*100) %>% round(1))

# saveRDS(plot.results,"C:/Users/n9232371/Documents/github/consult/finalwriteup/report_data/plot_finance.rds" )

ggplot(plot.results, aes(x= threshold, y= profit.ratio, colour = method)) + 
        geom_point(alpha = 0.1, shape = 16) + 
        stat_smooth() +
        scale_y_continuous(breaks = seq(0,110, by = 10)) + 
        scale_x_continuous(breaks = seq(0,1, by = .1)) + 
        geom_hline(yintercept = 100, colour = "darkgray", linetype = 2)

# why is it so shit? calc average prob for each iteration
mean.prob = apply(compile.results[['simp.log']], 1, function(x) mean(x[2:11])) 
analysis = data.frame(index = 1:length(mean.prob), mean.prob = mean.prob, 
                      bal = compile.results[['simp.log']]$balance.mlsto)
analysis %>% arrange(mean.prob) %>% head(10)
ggplot(analysis, aes(x = mean.prob, y = bal)) + geom_point()

Now need to check how mnay experiments to do.. power calc.

```{r}

#power calc - where should I do it from?

Check what happens when you increase the amount you invoiced by a certain percentage

```{r}

# what if you find for jobs over a certain threshold, increase fee by ?% still do all jobs
compile.results <- readRDS("C:/Users/n9232371/Documents/github/consult/finalwriteup/report_data/finance_results.rds")
####
# now have a suite of predictions - merge with original all10.blend$balance.mlsto
df.blend$index = rownames(df.blend) %>% as.numeric
# load balances
all6 <-read.csv('C:/Users/n9232371/OneDrive/shared files/Bligh Tanner/masters/data/all6.csv')
all6 <- all6 %>% select(mlsto, balance.mlsto, inv.mlsto, cost.mlsto) %>% distinct()
all6$mlsto <- as.character(all6$mlsto)
df.blend = left_join(df.blend, all6)
compile.results <- map(compile.results, function(x) {
        left_join(x, df.blend %>% select(index, balance.mlsto, b.rpdol, inv.mlsto, cost.mlsto))})

#create vector of profit points
threshold = seq(0, 1, by = .05)

# set up list of blank dataframes for profit calcs for each iteration: prof.results
colnames = names(compile.results[[1]])[grepl('iter', names(compile.results[[1]]))]
new.df = matrix(NA, nrow = length(threshold), ncol = ceiling(tot.tests/max.iter) + 1) %>% as.data.frame
colnames(new.df) <- c('threshold', colnames)
new.df$threshold = threshold 
prof.results = list(new.df, new.df, new.df, new.df, new.df)
names(prof.results) = names(compile.results)

prof.results <- map2(compile.results, prof.results, function(x,y) {

        for(n in 2:ncol(y)){

                for (p in 1:length(threshold)) {

                # sum the balance.mlsto of all jobs - you do them all
                # then sum a 10% increase in what you charge for all jobs above a certain threshold

                        prof <- x[, 'balance.mlsto']
                        increase <- x[x[,n] > y$threshold[p], 'inv.mlsto']*0.1

                        y[p, n] = sum(prof) + sum(increase)

        }
                return(y)
})

prof.results <- map2(prof.results, names(prof.results), function(x,y) {
        x$method = y
        return(x)})

assign(prof.results, "prof.results", envir = .GlobalEnv)

# gather results for plotting
plot.results = rbind_all(prof.results)
plot.results = gather(plot.results, key = iteration, value = profit, -method, - threshold)

ggplot(plot.results, aes(x= threshold, y= profit, colour = method)) + 
        geom_point(alpha = 0.1, shape = 16) + 
        stat_smooth()

# doesn't reveal much - better off doing a simple check once decided threshold.

# I like simp.log best.

simp.l <- compile.results[['simp.log']]
simp.l$mean.iter <- rowMeans(simp.l[,2:11])

# How many jobs are above threshold of 0.65?
simp.l %>% filter(mean.iter>0.65) %>% nrow
# 74 - not many

#how much was lost above 0.65?
initial = simp.l %>% filter(mean.iter>0.65) %>% .$balance.mlsto %>% sum
# $600000
# what if increased invoice amount by 0.1?
costs = simp.l %>% filter(mean.iter>0.65) %>% .$cost.mlsto %>% sum
simp.l %>% filter(mean.iter>0.65) %>% mutate(frac = inv.mlsto*0.27) %>% .$frac %>% sum

# would have to increase all the jobs which rated above 0.65 by 27% to break even.
# would have to increase all the jobs which rated above 0.65 by 46% to generate a 15% profit.
# this is probably impractical - hard to compete or argue for it. could also go for Hourly rate billing

profit.15 = costs*.15

### 2016-06-09 model_blend.rmd
report writing profit curve

@@

        #create model averaged column
        df = mutate(df, model.av = (log + rf +boost)/3)

@@

iter.auc(mult = 20, x.folds = 5, seedy = 100, df = all10.blend,
         methods = c("orig.log","log","average", "boost","rf", "simp.log","simp.boost"))
# iter.auc(mult = 10, x.folds = 10, seedy = 100, df = all10.blend,
#          methods = c("simp.log"))

# simp.log.notrans <- auc.list
# saveRDS(auc.list, "/Users/yam/Documents/github/consult/finalwriteup/report_data/blend_auc.rds")
auc.list <- readRDS("C:/Users/n9232371/Documents/github/consult/finalwriteup/report_data/blend_auc.rds")
# auc.list <- append(auc.list, simp.log.notrans)
# names(auc.list)[[8]] = "simp.log.notrans"
#don't worry about simp.log.notrans  - transformation of model predictions doesn't improve results at all

>>>>

formula.slog <- "f.rpdol ~ norm.log + norm.rf"

@@

        #create model averaged column
        df = mutate(df, model.av = (log + rf + boost)/3)

@@

iter.auc(mult = 20, x.folds = 5, seedy = 100, df = all10.blend,
         methods = c("orig.log","log","average", "boost","rf", "simp.log","simp.boost"))
# iter.auc(mult = 20, x.folds = 5, seedy = 100, df = all10.blend,
#          methods = c("simp.log", "average"))
# 
# no.boost <- auc.list
# saveRDS(auc.list, "/Users/yam/Documents/github/consult/finalwriteup/report_data/blend_auc.rds")
auc.list <- readRDS("C:/Users/n9232371/Documents/github/consult/finalwriteup/report_data/blend_auc.rds")
# auc.list <- append(auc.list, no.boost)
# names(auc.list)[c(8,9)] = c("simp.log.noboost","average.noboost")
#don't worry about simp.log.notrans  - transformation of model predictions doesn't improve results at all
